Evaluating impacts of Density Dependent processes on Blue whiting via MSE simulations

### 2023-03-13 Rmd/report mac.Rmd
updating reports

The OMs were also projected for the estimates of different estimates $F_{MSY}$, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**. **Figures 20 and 21** summarizes the yields and SSBs obtained for each Operating Model under the correct and mis-specified form of density dependence.

Conclusions

Estimating density-dependence from observations only possible for mass-at-age and maturity age. Density dependence in natural mortality can only be explored from theoretical considerations (because current assessment does not include age or time varying natural mortality).

The impact of including density dependence: major impact on reference points. Especially $F_{MSY}$ is much higher when including DD. However, the expected yields in equilibrium conditions are expected to be very similar. The main difference is in the initial step from the current $F_{MSY}$ to any new $F_{MSY}$ that would include DD, as the higher $F_{MSY}$ would be applied to a stock that is estimated without DD.

@@

**Figure 20** summarizes the yields and SSBs obtained for each Operating Model under the correct and misspecified form of density dependence and the associated reference points, **Figure 21** shows the Kobe plots of the final 2050 distributions of biomass and yield and **Figure 22** the Kobe pltos for F and yield.

>>>>

The OMs were also projected for the estimates of different estimates $F_{MSY}$, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**. **Figures 20-22** summarizes the yields, SSBs and Fs obtained for each Operating Model under the correct and mis-specified form of density dependence.

Estimating density-dependence from observations was only possible for mass-at-age and maturity age. Density dependence in natural mortality can only be explored from theoretical considerations (because current assessment does not include age or time varying natural mortality).

Including density dependence has a major impact on reference points. Especially $F_{MSY}$ is much higher when including DD. However, the expected yields in equilibrium conditions are expected to be very similar. The main difference is in the initial step from the current $F_{MSY}$ to any new $F_{MSY}$ that would include DD, as the higher $F_{MSY}$ would be applied to a stock that is estimated without DD.

@@

**Figure 20** summarizes the yields and SSBs obtained for each Operating Model under the correct and misspecified form of density dependence and the associated reference points, **Figure 21** shows the Kobe plots of the final 2050 distributions of biomass and yield and **Figure 22** the Kobe plots for F and yield.

### 2023-03-13 Rmd/report whb.Rmd
updating reports

---
output:
  html_document:
    reference_docx: ../report_template_v1.1.dotx
mathjax: TRUE
bibliography: refs.bib
---

@@

Version: 1.0 (`r format(Sys.time(), '%d %B %Y')`)

@@

@@

The OMs were also projected for the estimates of different estimates $F_{MSY}$, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**. **Figures 20 and 21** summarizes the yields and SSBs obtained for each Operating Model under the correct and mis-specified form of density dependence.

Conclusions

Estimating density-dependence from observations only possible for mass-at-age and maturity age. Density dependence in natural mortality can only be explored from theoretical considerations (because current assessment does not include age or time varying natural mortality).

The impact of including density dependence: major impact on reference points. Especially $F_{MSY}$ is much higher when including DD. However, the expected yields in equilibrium conditions are expected to be very similar. The main difference is in the initial step from the current $F_{MSY}$ to any new $F_{MSY}$ that would include DD, as the higher $F_{MSY}$ would be applied to a stock that is estimated without DD.

@@

A Generalized Additive Mixed Model (GAMM) was estimated to assess the relationship between mass-at-age and total biomass.

weights are predicted from two variables: a smoother of total biomass/mean(total biomass) and age. The resulting factor is shown below. There is no clear relationship between biomass and weight-at-age.

@@

The change in relative mass-at-age (i.e. observed - mean values) is plotted against total biomass, this allows the relationship between growth and total biomass to be compared across all ages. It is seen that there is no clear relationship between biomass and weight at age. A summary of the linear model is in the table below.

**???The trend of growth with biomass is -0.18, this was then used to model mass-at-age as:**

@@

**Deterministic projections at** $F_{MSY}$ for different scenarios: trends in weight, maturity and M-at-age

How can we understand the results of the deterministic runs? Below is a plot of the trends in stock mass-at-age, maturity-at-age and natural mortality-at-age, for the most relevant ages (1-6) and for the four different scenarios (Base, DDM, DDMM and DDMMM). The density dependent effect of maturity is strongest when both stock weight and maturity are density dependent. When natural mortality is also density dependent, this affects the stock size and hence also the density dependence for stock weight and maturity, leading to a smaller density dependent effects for those variables.

@@

**Stochastic projections at** $F_{MSY}$ for different scenarios

Using stochastic projections by including random processes for recruitment deviations and the $F_{MSY}$ values that are consistent with the different scenarios, leads to the overview below. SSB and total biomass are highest in the case of no density dependence. Catches are different in the first years of 'switching' to the new reference points, but afterwards, the catches for the different scenarios are highly comparable.

@@

Misspecification of Fmsy, i.e. using an Fmsy from one scenario and applying it to a different scenario has only limited impacts on the overall outcomes in terms of recruitment, biomass or yield. In the plot below, the scenarios for density dependence are shown in the columns and the Fmsy's are shown as the colors. For example a valid combination is the Base case with the red curves (Fmsy=0.33) or the M scenario with the green curve. Overall the misspecification does not results in noticeable differences in catch although they do have difference for biomass.

@@

For the four scenarios, we plotted the Kobe phase plot for 2050 with the associated density functions for B/Bmsy and Catch/MSY. The results confirm the results that equilibrium catch is not really affected by the DD assumptions of the Fmsy applied, but that biomass is sensitive to the assumptions.

@@

**Figure 20** summarizes the yields and SSBs obtained for each Operating Model under the correct and misspecified form of density dependence and the associated reference points, and **Figure 21** shows phase plots of the final 2050 distributions of biomass and yield. THese are for the Base Case, then density-dependence in mass, mass and maturity, and mass, maturity and natural mortality, projected for the four estimates of $F_{MSY}$. The correctly specified $F_{MSY}$ for each model is indicated by the red points and marginal densities.

The main diofferences are seen in $SSB/B_{MSY}$, if no density dependence is assumed then in the operating model the estimate of $F_{MSY}$ is the lowest, therefore projreting for the higher estimates of $F_{MSY}$ assuming density dependence results in an overfished stock. Yileld are the same for the two values of $F_{MSY}$ that assume no density-dependence on density dependence in mass-at-age only. In the other two cases yield is reduced but only by about 10%.

@@

Since the early 1980s, natural mortality ($M$) for `r mystkname` has been fixed within the assessment model at 0.2, for all ages and all assessment years.

M must, however, vary both by age and over time due to changes in abundance of predators, feeding conditions, disease etc. [@siegfried2009review]. Particularly any changes in `r mystkname` distribution and growth will affect natural mortality.

>>>>

---
title: "Evaluating impacts of Density Dependent processes on Blue whiting via MSE simulations"
author:
- name: L.T. Kell
  affiliation: Sea++
- name: M.A. Pastoors
  affiliation: MPFF
output:
  word_document:
    reference_docx: ../report_template_v1.1.dotx
mathjax: TRUE
bibliography: refs.bib
---

@@

Version 1.0 (`r format(Sys.time(), '%d %B %Y')`)

&nbsp;

@@

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

@@

The OMs were also projected for the estimates of different estimates $F_{MSY}$, with stochasticty in the recruitment deviates. This allows the impact of misspecification of density dependence to be evaluated, i.e. if $F_{MSY}$ is underestimated then yield will be less than MSY, while if $F_{MSY}$ is overestimated yield will be lost due to overfishing. Stochastic projections of the four Operating Models for the four estimates of $F_{MSY}$ are shown in **Figures 16-19**. **Figures 20-22** summarizes the yields, SSBs and Fs obtained for each Operating Model under the correct and mis-specified form of density dependence.

Estimating density-dependence from observations was not really possible for `r mystkname`. Therefore, density dependence in mass at age, maturity at age and natural mortality has only be explored from theoretical considerations in this report.

Including density dependence has a major impact on reference points. Especially $F_{MSY}$ is much higher when including DD. However, the expected yields in equilibrium conditions are expected to be very similar. The main difference is in the initial step from the current $F_{MSY}$ to any new $F_{MSY}$ that would include DD, as the higher $F_{MSY}$ would be applied to a stock that is estimated without DD.

@@

A Generalized Additive Mixed Model (GAMM) was estimated to assess the relationship between mass-at-age and total biomass. Weights are predicted from two variables: a smoother of total biomass/mean(total biomass) and age. The resulting factor is shown below. There is no clear relationship between biomass and weight-at-age.

@@

The change in relative mass-at-age (i.e. observed - mean values) is plotted against total biomass, this allows the relationship between growth and total biomass to be compared across all ages. Here, we plotted both the GAMM model and the Linear Model variants. It is seen that there is still some weak information on density dependence in the Linear Model (slope = beta = -0.11). A summary of the linear model is in the table below.

@@

**Deterministic projections at $F_{MSY}$ for different scenarios: trends in weight, maturity and M-at-age**

How can we understand the results of the deterministic runs? Below is a plot of the trends in stock mass-at-age, maturity-at-age and natural mortality-at-age, for the most relevant ages (1-6) and for the four different scenarios (Base, DDM, DDMM and DDMMM). The density dependent effect of maturity is strongest when both stock weight and maturity are density dependent. When natural mortality is also density dependent, this has only minor impacts.

**@Laurie: why would DD in natural mortality have such a minimal impact? Did you see that the M levels overall are lower than the assumed M in the assessment (=0.2)**

@@

**Stochastic projections at $F_{MSY}$ for different scenarios**

Using stochastic projections by including random processes for recruitment deviations and the $F_{MSY}$ values that are consistent with the different scenarios, leads to the overview below. SSB and total biomass are highest in the case of no density dependence. Biomass tends to be lower when density dependent processes are taken into account. Catches are different in the first years of 'switching' to the new reference points, but afterwards, the catches for the different scenarios are highly comparable.

@@

Misspecification of Fmsy, i.e. using an Fmsy from one scenario and applying it to a different scenario has only limited impacts on the overall outcomes in terms of recruitment, biomass or yield. In the plot below, the scenarios for density dependence are shown in the columns and the Fmsy's are shown as the colors. For example a valid combination is the Base case with the red curves (Fmsy=0.27) or the M scenario with the green curve. Overall the misspecification does not results in noticeable differences in catch although they do have important difference for biomass.

@@

For the four scenarios, we plotted the Kobe phase plot for 2050 with the associated density functions for B/Bmsy and Catch/MSY. The results confirm the results that equilibrium catch is not really affected by the DD assumptions of the Fmsy applied, but that biomass is sensitive to the assumptions. Application of density dependent reference points when density dependency would not exist, is expected to lead to lower equilibrium biomass levels.

@@

**Figure 20** summarizes the yields and SSBs obtained for each Operating Model under the correct and misspecified form of density dependence and the associated reference points, **Figure 21** shows the Kobe plots of the final 2050 distributions of biomass and yield and **Figure 22** the Kobe plots for F and yield. These are for the Base Case, then density-dependence in mass, mass and maturity, and mass, maturity and natural mortality, projected for the four estimates of $F_{MSY}$. The correctly specified $F_{MSY}$ for each model is indicated by the red points and marginal densities.

The main differences are seen in $SSB/B_{MSY}$, if no density dependence is assumed then in the operating model the estimate of $F_{MSY}$ is the lowest, therefore projecting for the higher estimates of $F_{MSY}$ assuming density dependence results in an overfished stock. Yield are the same for the two values of $F_{MSY}$ that assume no density-dependence on density dependence in mass-at-age only. In the other two cases yield is reduced by about 10%.

@@

Since the early 1980s, natural mortality ($M$) for `r mystkname` has been fixed within the assessment model at 0.2, for all ages and all assessment years.Natural mortality must, however, vary both by age and over time due to changes in abundance of predators, feeding conditions, disease etc. [@siegfried2009review]. Particularly any changes in `r mystkname` distribution and growth will affect natural mortality.
